---
title: "LAB 9 - Candy Mini-Project"
author: "Karla Dominguez Rivera"
format: pdf
editor: visual
---

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names = 1)
head(candy)
```

>***2.1 What is in the dataset? ***


>**Q1- How many different candy types are in this dateset? **

There is 85 different types of candy in this datset.

```{r}
nrow(candy)
```


>**Q2- How many fruity canfy types are there in the dataset?**

There is 38 fruity candy types in the dataset.

```{r}
sum(candy$fruity)
```



>***2.2 What is your favorite candy?***


```{r}
candy["Twix", ]$winpercent
```


```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Twix") |> 
  select(winpercent)
```


>**Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value? **

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Snickers") |> 
  select(winpercent)
```


>**Q4. What is the winpercent value for “Kit Kat”? **

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Kit Kat") |> 
  select(winpercent)
```


>**Q5. What is the winpercent value for “Tootsie Roll Snack Bars”? **

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Tootsie Roll Snack Bars") |> 
  select(winpercent)
```

```{r}
library("skimr")
skim(candy)
```
>**Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?**

Yes winpercent is on a different scale that is roughly 0–100 compared to most other columns, which are binary 0/1 variables.

>**Q7. What do you think a zero and one represent for the candy$chocolate column?**

 In the chocolate column, 1 means the candy contains chocolate and 0 means it does not.

```{r}
table(candy$chocolate)
```

>**Q8. Plot a histogram of winpercent values using both base R an ggplot2.**

```{r}
# Base R histogram of winpercent
hist(candy$winpercent,
     main = "Histogram of Candy Win Percent",
     xlab = "Win Percent",
     col = "lightblue",)
```



```{r}
library(ggplot2)

# ggplot2 histogram of winpercent
ggplot(candy, aes(x = winpercent)) +
  geom_histogram(binwidth = 5, fill = "lightblue", color = "black") +
  labs(title = "Histogram of Candy Win Percent",
       x = "Win Percent",
       y = "Count")
```

>**Q9. Is the distribution of winpercent values symmetrical?**

No, the distribution is slightly skewed not perfectly symmetrical.


>**Q10. Is the center of the distribution above or below 50%?**

```{r}
# Calculate mean winpercent
mean(candy$winpercent)
```
 The center is below 50 %

>**Q11. On average is chocolate candy higher or lower ranked than fruit candy?**

```{r}
# Mean winpercent for chocolate vs fruity candies
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])
```

Chocolate candy is ranked higher on average than fruity candy.

>**Q12. Is this difference statistically significant?**

```{r}
# Perform t-test comparing chocolate vs fruity candies
t.test(candy$winpercent[as.logical(candy$chocolate)],
       candy$winpercent[as.logical(candy$fruity)])
```
Yes, the difference is statistically significant with a very small p-value of ≈ 2.87e-08.

```{r}
library(dplyr)

# Bottom 5 (least liked candies)
candy |> 
  arrange(winpercent) |>   # sort from lowest to highest
  head(5)

# Top 5 (most liked candies)
candy |> 
  arrange(desc(winpercent)) |>  # sort from highest to lowest
  head(5)
```
```{r}
# Bottom 5
head(candy[order(candy$winpercent), ], 5)

# Top 5
head(candy[order(-candy$winpercent), ], 5)
```


>**Q13. What are the five least liked candy types in this set?**

The five least liked candy ypes in this set are Nik L Nip,Boston Baked Beans
Chiclets,Super Bubble, and Jawbusters.

>**Q14. What are the top 5 all time favorite candy types out of this set?**

The five most liked candy ypes in this set are Reese’s Peanut Butter Cup
Reese’s Miniatures, Twix, Kit Kat, and Snickers

>**Q15. Make a first barplot of candy ranking based on winpercent values.** 

```{r, fig.height=12, fig.width=7.5}
library(ggplot2)

# Basic barplot of winpercent vs candy names
ggplot(candy) + 
  aes(x = winpercent, y = rownames(candy)) +  # map winpercent to x-axis, candy names to y-axis
  geom_col()   # use geom_col() to create bars with actual values
```
>**Q16 This is quite ugly, use the reorder() function to get the bars sorted by winpercent?**

```{r}
library(ggplot2)

# Barplot with candies sorted by winpercent
ggplot(candy) + 
  aes(x = winpercent, 
      y = reorder(rownames(candy), winpercent)) +  # reorder candy names by winpercent
  geom_col() +
  labs(title = "Candy Rankings by Win Percent",
       x = "Win Percent",
       y = "Candy Type")
```
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r, fig.height=12, fig.width=7.5}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) +
  theme(axis.text = element_text(size = 12))
```
>**Q17. What is the worst ranked chocolate candy?**

The worst ranked chocolae candy bar was Nik L Nip 

>**Q18. What is the best ranked fruity candy?**

The nest ranked fruity bar was Reeses penut butter cup.


```{r}
library(ggrepel)

# Plot of winpercent vs pricepercent
ggplot(candy) +
  aes(x = pricepercent, y = winpercent, label = rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size = 3.3, max.overlaps = 5)
```
>**Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?*

The candy with the highest winpercent relative to its price is Reese’s Miniatures

```{r}
# Create value score: popularity per price
candy$value_score <- candy$winpercent / candy$pricepercent

# Find the best "bang for your buck"
candy[order(-candy$value_score), c("winpercent", "pricepercent")]
```


>**Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?**

the top 5 most expensive candies are Nik L Nip, Nestle Smarties, Ring Pop
Hershey’s Krackel and Hershey’s Milk Chocolate.

```{r}
# Order by most expensive first
ord <- order(candy$pricepercent, decreasing = TRUE)

# Top 5 most expensive candies
top5_expensive <- candy[ord, c("pricepercent", "winpercent")][1:5, ]
top5_expensive
```

>**Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().**

```{r, fig.height=12, fig.width=7.5}
library(ggplot2)

# Basic barplot of pricepercent
ggplot(candy) +
  aes(x = pricepercent, y = rownames(candy)) +
  geom_col()
```
```{r,fig.height=12, fig.width=7.5}
# Lollipop chart of pricepercent
ggplot(candy) +
  aes(x = pricepercent,
      y = reorder(rownames(candy), pricepercent)) +
  
  # stems of the lollipop
  geom_segment(aes(
    x = 0,
    xend = pricepercent,
    yend = reorder(rownames(candy), pricepercent)
  ), col = "gray40") +
  
  # dots of the lollipop
  geom_point()
```

```{r}
library(corrplot)

```

```{r}
candy_BK <- candy
candy <- candy %>% dplyr::select(-value_score)

cij <- cor(candy)
corrplot(cij)
```
>**Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?**

Chocolate and fruity are the most strongly anti-correlated variables in the dataset.

>**Q23. Similarly, what two variables are most positively correlated?**

Chocolate and winpercent are the most strongly positively correlated variables in the dataset.

```{r}
# Run PCA with scaling
pca <- prcomp(candy, scale = TRUE)

# Summary of PCA results
summary(pca)
plot(pca$x[,1:2])
```



```{r, fig.height=5, fig.width=6}
# Run PCA
pca <- prcomp(candy, scale = TRUE)

# Summary of PCA results
summary(pca)

# Plot PC1 vs PC2 with custom axis limits
plot(pca$x[,1:2],
     xlim = c(-4, 2),   # x-axis range
     ylim = c(-3, 2),   # y-axis range
     xlab = "PC1",
     ylab = "PC2")
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
# Make a new data-frame with PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

# PCA scatter plot (PC1 vs PC2)
p <- ggplot(my_data) + 
        aes(x = PC1, y = PC2, 
            size = winpercent / 100,  
            text = rownames(my_data),
            label = rownames(my_data)) +
        geom_point(col =my_cols)
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
ggplotly(p)
```

>**Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?**

PC1 is strongly driven in the positive direction by chocolate (and related bar-type variables), which matches what we previously saw in the correlation analysis where chocolate was positively associated with winpercent and negatively associated with fruity candy.

```{r}
library(ggplot2)

# Put PCA rotation into a data frame with variable names
loadings <- as.data.frame(pca$rotation)
loadings$variable <- rownames(loadings)

# Bar plot of PC1 loadings
ggplot(loadings) +
  aes(x = reorder(variable, PC1), y = PC1) +
  geom_col() +
  coord_flip()
```

>**Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?**

 A “winning” candy is typically chocolate-based, often in bar form, and not fruity, and this conclusion is consistently supported across visualization, correlation analysis, and PCA, which all show chocolate as the higher popularity.
