---
title: "Lab 8: Unsupervised Learning Mini-Project"
format: pdf
author: Karla Dominguez Rivera
editor: visual
---

>**PART 1 Preparing the data **


```{r}
# Save your input data file into your Project directory
fna.data <- "WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data, row.names=1)
```


```{r}
head(wisc.df, 4)
```


```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
```

```{r}
diagnosis <- factor(wisc.df$diagnosis)
```

>**Q1. How many observations are in this dataset?**

These is 569 observations.
```{r}
nrow(wisc.df)
```


>**Q2. How many of the observations have a malignant diagnosis?**

212 observations have maligant diagnostics. 

```{r}
table(diagnosis)
```


>**Q3.How many variables/features in the data are suffixed with _mean?**

There is 10 variable features in the data that are suffixed with _mean.
```{r}
sum(grepl("_mean", names(wisc.df)))
```


>***PART 3. Principal Component Analysis***

```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)


```


```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale. = TRUE)
```

```{r}
summary(wisc.pr)
```

>**Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?**

The first principal component (PC1) captures about 44.27% of the total variance in the data.

>**Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?**

```{r}
pr.var <- wisc.pr$sdev^2
pve <- pr.var / sum(pr.var)
cumulative_pve <- cumsum(pve)
which(cumulative_pve >= 0.70)[1]
```

You need 3 principal components to explain at least 70% of the variance.

>**Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?**

```{r}
pr.var <- wisc.pr$sdev^2
pve <- pr.var / sum(pr.var)
cumulative_pve <- cumsum(pve)
which(cumulative_pve >= 0.0)[1]
```


You need 7 principal components to explain at least 90% of the variance.

>**Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?**

The biplot is very cluttered and difficult to interpret with many data points and overlapping each other.It is hard to clearly see patterns or groupings in the data making the plot not very useful for understanding the PCA results.

```{r}
biplot(wisc.pr)
```

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col = diagnosis) +
  geom_point()
```

>**Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?**

The plot of PC1 vs PC3 still shows separation between benign and malignant cases, but it is not as clear as PC1 vs PC2.PC1 appears to be the strongest component for distinguishing the groups.PC3 adds some variation, but it does not separate the classes as well as PC2 does in combination with PC1.

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col = diagnosis) +
  geom_point()

```


```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```


```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

```{r}
## ggplot based graph
#install.packages("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

>>**Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?**

The loading value for concave.points_mean on PC1 is large and positive around 0.26.Yes, there are a few features with even larger contributions to PC1.

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```

> ***PART 4:  Hierarchical clustering ***

```{r}
data.scaled <- scale(wisc.data)
```


```{r}
data.dist <- dist(data.scaled)
```

```{r}
wisc.hclust <- hclust(data.dist, method = "complete")
```

>**Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters? **

The height where the clustering produces 4 clusters is approximately 20.

```{r}
plot(wisc.hclust)
abline(h = 20, col = "red", lty = 2)
```

>**Q11. OPTIONAL: Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 6? How do you judge the quality of your result in each case?**

Yes, you can try different numbers of clusters.Often, 2 clusters already gives a very strong separation because the data naturally splits into benign vs malignant

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
```


>**Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.**

the medthod ward.D2 gives the best results because it creates the most compact and well-separated clusters, which most closely match the true benign vs malignant labels in the dataset.
```{r}
wic.hclust.complete <- hclust(data.dist, method= "complete")
wic.hclust.average <- hclust(data.dist, method= "average")
wic.hclust.single <- hclust(data.dist, method= "single")
wic.hclust.ward <- hclust(data.dist, method= "ward")
View(wisc.df)
```


> ***PART 5: Combining methods***


```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method = "ward.D2")
```


```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method = "ward.D2")
plot(wisc.hclust)
```

```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method = "ward.D2")
plot(wisc.pr.hclust)
```



```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```


```{r}
table(grps, diagnosis)
```

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point(col=grps)


```


optional : 
```{r}
library(rgl)
plot3d(wisc.pr$x[,1:3], xlab="PC 1", ylab="PC 2", zlab="PC 3", cex=1.5, size=1,
       type="s", col=grps)
```


```{r}
## Use the distance along the first 7 PCs for clustering i.e. wisc.pr$x[, 1:7]
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method = "ward.D2")

```




```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k = 2)
```


>**Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?**

```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```

>**Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.**

The hierarchical clustering model without PCA does not separate the diagnoses very well. Several clusters contain a mixture of benign and malignant cases instead of clean separation.

```{r}
table(wisc.hclust.clusters, diagnosis)
```

>***PART 6 : Sensitivity/Specificity***


>**Q15. OPTIONAL: Which of your analysis procedures resulted in a clustering model
with the best specificity? How about sensitivity?**

The PCA-based hierarchical clustering model gives the best performance overall.
It has the highest sensitivity because it correctly identifies most malignant cases.

>***PART 7 :Prediction**  

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

>**Q16. Which of these new patients should we prioritize for follow up based on your results? **

Prioritize Patient 1 for follow-up because Patient 1’s PCA scores place them closer to the cluster associated with malignant tumors, which suggests a higher risk.Patient 2 appears to fall in the region more associated with benign cases, so they are lower priority.
